When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-20 doc/paper/vpd_et_paper.tex
makes me think, should we set our own uWUE so we can include EBF?

@@

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and relative humidity drying ** RH drying? \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand for evapotranspirated water \citep{Penman_1948, Monteith_1965}, but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}. These  conflicting responses, increased atmospheric demand and higher stomatal stress, leads to two possible responses of evapotraspiration (ET) to changes in VPD. The first, hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion \citep{Swann_2016}. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. Therefore, an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.

@@

We can use intuition about plant water conservation strategy to hypothesize about ET response to changes in VPD. Plants that evolved to conserve water, such as arid shrubs, should be more likely to reduce ET with increasing VPD, and plants that have evolved or have been engineered to care little about water, such as crops, will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale, there are limits to plant water conservation strategies. As atmospheric demand for water (VPD) increases further, ecosystems should begin to reach their water conservation limits and might not be able to entirely limit the ET flux to the atmosphere. At this stage any further increase in VPD will most likely drive a (limited) increase in ET, because the increase in atmospheric demand for water overwhelms the limited plant response to conserve water.

The objective of the present manuscript is thus to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to drive intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types. Section 2 describes the data used. Section 3 derives the framework. Section 4 presents results and comparison to observations. Section 5 discusses conclusions.

\section{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015 database, including all sites with more than four years of data (data available at https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/). Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used for water use efficiency in Section \ref{methods} are only published for five plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 56 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now}.

@@

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining an optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with an empirical approach, which describes the dependence of $g_s$ to VPD. This resulted in the following model for leaf-scale stomatal conductance was:

\begin{linenomath*}
  \begin{equation}
  g_{l-s} = g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_s}
  \end{equation}
\end{linenomath*}
Where $g_1$ is a leaf-scale ``slope'' parameter, A is the net CO$_2$ assimilation rate, and $c_s$ is the atmospheric CO$_2$ concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope parameter ($g_1$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
  g_1 = \sqrt{\frac{3 \, \Gamma \, \lambda}{1.6}},\footnote{Note the above expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

where $\Gamma$ is the CO$_2$ compensation point for photosynthesis (without dark respiration), and $\lambda$ is the marginal cost of leaf carbon. So, $g_1$ is a leaf-scale term reflecting the tradeoff of water for carbon uptake and growth. The higher $g_1$, the more open are the stomata and the more they release water in exchange for carbon.

@@

While Equation 3 can be used in PM (Equation 1), it will make analytical work with the function intractable because $A$, net CO$_2$ assimilation, is functionally related to ET itself. To remove the dependence of ET on $A$ we use a fundamental semi-empirical results of \citet{Zhou_2015}, who showed that the underlying Water Use Efficiency $uWUE$:

\begin{linenomath*}
  \begin{equation}
    \label{uwue}
uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
  \end{equation}
\end{linenomath*}
is relatively constant across time and moisture conditions within a plant functional type (CITE ALSO CHAGJIE ** ??), in non-extreme drought conditions. If, following \citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect cuticular and epidermal losses - a reasonable assumption except in very dry conditions), we can use uWUE to remove the $A$ dependence from $g_s$ in a way that makes the PM equation analytically tractable:

\begin{linenomath*}
  \begin{equation}
  g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{VPD}}
  \end{equation}
\end{linenomath*}

Note that $uWUE$ is fit on the ecosystem scale as in \citet{Zhou_2015} so GPP in \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancellation of LAI in addition to uWUE in Equation \ref{medlyn}. Plugging Equation 5 into Equation 1 and rearranging gives a new expression for PM ET directly as a function of VPD with no $A$ dependence:

@@

With Equation \ref{et}, ET is explicitly a function of environmental variables and two plant-specific constants, the slope parameter ($g_1$), and $uWUE$, both reflecting water conservation strategy. The slope parameter is a leaf-scale term related to the willingness of stomata to trade water for CO$_2$ and to keep stomata open. $uWUE$ is semi-empirical ecosystem-scale constant related to how WUE changes with VPD (specifically $VPD^{-1/2}$). In the eddy-covariance FLUXNET2015 data described in Section \ref{data}, every term in Equation \ref{et} is observed, except for the two parameters $g_1$, and $uWUE$, which have however been measured for the PFTs considered here \citep{Zhou_2015, Lin_2015}. However, eddy-covariance data are inherently noisy so that we include a measure of uncertainty in our analysis. To account for both mean deviations of $uWUE$ and uncertainty, we introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the psychometric constant ($\gamma$) are relatively constant over the dataset considered here so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions  (ZHOU ET AL, LIN ET AL ** ??)\explain{Pierre, which Zhou and Lin citations were you referring to that show that uWUE changes with very dry soil conditions?}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from a decrease in ET, I.e. the threshold $VPD_{crit}$ such as the derivative vanishes $\frac{\partial \; ET}{\partial \; VPD} = 0$:

\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + \sqrt{\frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE}\left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}

We note that $VPD_{crit}$ mostly depends on the PFT parameters $uWUE$ and $g_1$ but only vary weakly with climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation \citep{De_2013, Nikinmaa_2013, Ryan_2014}.

@@

\section{Conclusions}

>>>>

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}

@@

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and relative humidity drying ** RH drying? \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand for evapotranspirated water \citep{Penman_1948, Monteith_1965}, but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}. The conflicting effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotraspiration (ET) respondes to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion \citep{Swann_2016}. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

We can use intuition about plant water conservation strategy to hypothesize about ET response to changes in VPD. Plants that evolved to conserve water, such as arid shrubs, should be more likely to reduce ET with increasing VPD, and plants that have evolved or have been engineered to care little about water, such as crops, will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale, there are limits to plant water conservation strategies. As atmospheric demand for water (VPD) increases further, ecosystems should begin to reach their water conservation limits and might not be able to entirely limit ET flux to the atmosphere. At this stage any further increase in VPD will most likely drive a (limited) increase in ET, because the increase in atmospheric demand for water overwhelms the limited plant response to conserve water.

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to drive intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types. Section 2 describes the data used and derives the framework. Section 4 presents results and comparison to observations, and  Section 5 discusses the conclusions.

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015 database, including all sites with more than four years of data (data available at https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/). Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used for water use efficiency in the Methods section (Section \ref{methods}) are only published for five plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 56 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now}.

@@

\subsection{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining an optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with an empirical approach, which describes the dependence of $g_s$ to VPD. This resulted in the following model for leaf-scale stomatal conductance was:

\begin{linenomath*}
  \begin{equation}
  g_{l-s} = g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_s}
  \end{equation}
\end{linenomath*}
Where $g_1$ is a leaf-scale ``slope'' parameter, A is the net CO$_2$ assimilation rate, and $c_s$ is the atmospheric CO$_2$ concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope parameter ($g_1$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
  g_1 = \sqrt{\frac{3 \, \Gamma^* \, \lambda}{1.6}},\footnote{Note the above expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis (without dark respiration), and $\lambda$ is the marginal water cost of leaf carbon. So, $g_1$ is a leaf-scale term reflecting the tradeoff of water for carbon uptake and growth. The higher $g_1$, the more open the stomata and the more they release water in exchange for carbon.

@@

While Equation 3 can be used in PM (Equation 1), it will make analytical work with the function intractable because $A$, net CO$_2$ assimilation, is functionally related to ET itself. To remove the dependence of ET on $A$ we use fundamental semi-empirical results of \citet{Zhou_2015}, who showed that the underlying Water Use Efficiency ($uWUE$):

\begin{linenomath*}
  \begin{equation}
    uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*}
is relatively constant across time and moisture conditions within a plant functional type (CITE ALSO CHAGJIE ** ??), in non-extreme drought conditions. If, following \citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect cuticular and epidermal losses - a reasonable assumption except in very dry conditions), we can use uWUE to remove the $A$ dependence from $g_s$ in a way that makes the PM equation analytically tractable:

\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{VPD}}
    \label{new_g_s}
  \end{equation}
\end{linenomath*}

Note that $uWUE$ is fit on the ecosystem scale as in \citet{Zhou_2015} so GPP in Equation \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancellation of LAI with the addition of uWUE in Equation \ref{medlyn}. Plugging Equation \ref{new_g_s} into Equation \ref{orig_pen} and rearranging gives a new expression for PM ET directly as a function of VPD with no $A$ dependence:

@@

With Equation \ref{et}, ET is explicitly a function of environmental variables and two plant-specific constants, the slope parameter ($g_1$), and $uWUE$, both reflecting water conservation strategy. The slope parameter is a leaf-scale term related to the willingness of stomata to trade water for CO$_2$ and to keep stomata open. $uWUE$ is semi-empirical ecosystem-scale constant related to how WUE changes with VPD (specifically $VPD^{-1/2}$). Is is also roughly proportional to physical constants:

\[uWUE \appropto \sqrt{\frac{c_s - \Gamma}{1.6 \lambda}},\]

where $\Gamma$ is the CO$_2$ compensation point \citep{Zhou_2014}. So uWUE is related to atmospheric CO$_2$ concentration and compensation point, and is inversely proportional to the marginal water cost of leaf carbon. Therefore, generally speaking, plants that are very willing to trade water for new growth will have lower $uWUE$ and higher $g_1$, and will be less water conservative (more likely to increase ET with increasing VPD).

Given the eddy-covariance FLUXNET2015 data described in Section \ref{data}, every term in Equation \ref{et} is observed, except for the two parameters $g_1$, and $uWUE$, which have however been measured for the PFTs considered here \citep{Zhou_2015, Lin_2015}. However, eddy-covariance data are inherently noisy so we include a measure of uncertainty in our analysis. To account for both mean deviations of $uWUE$ and uncertainty, we introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the psychometric constant ($\gamma$) are relatively constant over the dataset considered here so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions  (ZHOU ET AL, LIN ET AL ** ??)\explain{Pierre, which Zhou and Lin citations were you referring to that show that uWUE changes with very dry soil conditions?}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from a decrease in ET, i.e. the threshold $VPD_{crit}$ such as the derivative vanishes $\frac{\partial \; ET}{\partial \; VPD} = 0$:
\small
\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + \sqrt{\frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE}\left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}
\normalsize
We note that $VPD_{crit}$ mostly depends on the PFT parameters $uWUE$ and $g_1$ but only vary weakly with climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation \citep{De_2013, Nikinmaa_2013, Ryan_2014}.

@@

\section{Discussion}
